---
title: "Sorting interpretation"
date: 2020-11-27
categories: ["R"]
tags: ["sort", "statistics", "plot"]
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
```

# Introduction

Sorting is non-parametric approach. Using sorting we can analyze the relationship between x and y. The advantage of sorting are 1) the ability to capture non-linearity, 2) easy to implement, 3) clear interpretation of the result. The disadvantage of sorting is difficulties to control for many variables.

The intuition behind sorting is similar with the basic in statistics (for example, nearest K neighborhood), where we can calculate $E[y|x = X]$ by using some data point near $x = X$.

Sorting is widely used in finance research since it reflects the portfolio return.

# Simulated data

```{r}
library(gridExtra)
library(ggplot2)
library(dplyr)

```


$test\_score\_n=\beta_1 ×education+ \beta_2 × effort+ \beta_3 ×education×effort + \epsilon$

```{r}


# simulate data
N_obs = 1000
set.seed(1)
df_master <- data.frame(education = runif(n = N_obs, min = 0, max = 100),
                        effort = runif(n = N_obs, min = 0, max = 100) )

# generate test score with noise
df_master <- df_master %>% 
  mutate(test_score = 2*education + 4*effort + 0.2*education*effort,
         # test_score_n is test score with noise. 
         test_score_n = test_score + rnorm(n = N_obs, mean = 0, sd = sd(test_score)/10 ) )  

# pairs
# pairs(df_master %>% select(-c(test_score)))

library(GGally)
ggpairs(df_master %>% select(-test_score))

```

# Sorting on y

Conditional on high test score, what is the average of education.

$E[education | test\_score\_n = group\ 5]$

The result of sort is as expected when we check the plot region. The group 1 (group 5) has the highest (lowest) standard deviation and lowest (highest) mean in X dimension.

```{r}
##### Sorting
df_sort <- df_master %>%
  # assign port based on y
  mutate(port = ntile(test_score_n, 5) ) %>%
  # calculate mean and sd of x
  group_by(port) %>%
  summarise(mean_x = mean(education),
            sd_x = sd(education) ) %>%
  ungroup()

# ggplot(data = df_sort, aes(x = port, y = mean_x, ymin = mean_x - sd_x,
#                            ymax = mean_x + sd_x)) +
#   geom_pointrange()

ggplot(data = df_sort, 
       aes(x = mean_x, y = port, xmin = mean_x - sd_x,
                           xmax = mean_x + sd_x)) +
  geom_pointrange()

##### plot by region
df_line <- df_master %>%
  # assign port based on y
  mutate(port = ntile(test_score_n, 5) )  %>%
  group_by(port) %>%
  summarise(test_score_n_min = min(test_score_n),
            test_score_n_max = max(test_score_n),
            test_score_n_mean = mean(test_score_n),
            ) %>%
  ungroup()

#Breaks for background rectangles
# ref for different color background: https://stackoverflow.com/questions/9968975/make-the-background-of-a-graph-different-colours-in-different-regions
rects <- data.frame(ystart = df_line$test_score_n_min, 
                    yend = df_line$test_score_n_max, col = as.factor(1:5))

ggplot() +
  geom_rect(data = rects, 
            aes(xmin = -Inf, xmax = Inf, ymin = ystart, ymax = yend,
                              fill = col), alpha = 0.4) +
  geom_point(data = df_master %>% 
         select(education, test_score_n ),
       aes(x = education, y = test_score_n)  ) 




```

# Sorting on x

Conditional on high education, what is the average of test score. The result is as expected as we see the pairs plot between education and test_score_n.

$E[test\_score\_n  | education  = group\ 5]$

The result of sort is as expected when we check the plot region. The group 1 (group 5) has the lowest (highest) standard deviation and lowest (highest) mean in Y dimension.

```{r}
##### Sorting
df_sort <- df_master %>%
  # assign port based on x
  mutate(port = ntile(education, 5) ) %>%
  # calculate mean and sd of y
  group_by(port) %>%
  summarise(mean_y = mean(test_score_n),
            sd_y = sd(test_score_n) ) %>%
  ungroup()

ggplot(data = df_sort, aes(x = port, y = mean_y, ymin = mean_y - sd_y,
                           ymax = mean_y + sd_y)) +
  geom_pointrange()

##### plot by region
df_line <- df_master %>%
  # assign port based on y
  mutate(port = ntile(education, 5) )  %>%
  group_by(port) %>%
  summarise(education_min = min(education),
            education_max = max(education),
            education_mean = mean(education),
            ) %>%
  ungroup()

#Breaks for background rectangles
# ref for different color background: https://stackoverflow.com/questions/9968975/make-the-background-of-a-graph-different-colours-in-different-regions
rects <- data.frame(xstart = df_line$education_min, 
                    xend = df_line$education_max, col = as.factor(1:5))

ggplot() +
  geom_rect(data = rects, 
            aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf,
                              fill = col), alpha = 0.4) +
  geom_point(data = df_master %>% 
         select(education, test_score_n ),
       aes(x = education, y = test_score_n)  ) 





```




