---
title: "Interaction effect in linear regession"
date: 2020-11-20
categories: ["R"]
tags: ["regression", "interaction effect", "plot"]
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
```

# Interpretation

Interaction effect is important when we analyzing data. Suppose there is positive relationship between *education* and *effort* on *test result*. Furthermore, there is synergy effect between *education* and *effort*. Given high *education*, a unit increase in effort can lead to higher *test result* compared to those who has low *education*.

$test\_score=\beta_1 ×education+ \beta_2 × effort+ \beta_3 ×education×effort + \epsilon$

<!-- $\frac{\partial \  test\_score}{\partial \ education} =\beta_1 + \beta_3 ×effort $ -->

Taking derivative wrt education:

$\partial \  test\_score / \partial \ education =\beta_1 + \beta_3 ×effort$


Hence, a unit change in education is related with $\beta_1 + \beta_3 ×effort$ change in test_score. In other words, the effect of increase in education depends on effort level.

# Visualizing

In the figure below, I simulated 3 different case.

- Case 1: $test\_score = 2 \times education + 4 \times effort + 2 \times education \times effort$
- Case 2: $test\_score = 2 \times education + 4 \times effort + 0.2 \times education \times effort$
- Case 3: $test\_score = 2 \times education + 4 \times effort + 0 \times education \times effort$

And the resulting test_score is adjusted such that ranging from 0 to 100. FUrthermore, I added noise in each case. As shown in the figure, ignoring the interaction effect will underfit the data.

**The effect of interaction shown in heatmap**. There is non-linearity for the interaction or synergy effect.

```{r}
library(gridExtra)
library(ggplot2)
library(dplyr)

#############################################
##### P1
df_grid <- expand.grid(education=seq(0,100, by=1), effort=seq(0,100,by=1))
df_grid <- df_grid %>% 
  mutate(test_score = 2*education + 4*effort + 2*education*effort,
         # test_score_n is test score with noise. 
         test_score_n = test_score + rnorm(n = nrow(df_grid), mean = 0, sd = sd(test_score)/10 ) ) %>%
  mutate(test_score_n = ntile(test_score_n, 100))

p1 <- ggplot(df_grid, aes(x = education, y = effort, z = test_score_n)) +
  # stat_contour(geom = "polygon", aes(fill = ..level..)) +
  geom_tile(aes(fill = test_score_n)) +
  scale_fill_gradient(low = "green", high = "red")+ 
  stat_contour(bins = 5) +
  xlab("education") +
  ylab("effort") +
  guides(fill = guide_colorbar(title = "lab")) 

##### P2
df_grid <- expand.grid(education=seq(0,100, by=1), effort=seq(0,100,by=1))
df_grid <- df_grid %>% 
  mutate(test_score = 2*education + 4*effort + 0.2*education*effort,
         # test_score_n is test score with noise. 
         test_score_n = test_score + rnorm(n = nrow(df_grid), mean = 0, sd = sd(test_score)/10 ) )  %>%
  mutate(test_score_n = ntile(test_score_n, 100))


p2 <- ggplot(df_grid, aes(x = education, y = effort, z = test_score_n)) +
  # stat_contour(geom = "polygon", aes(fill = ..level..)) +
  geom_tile(aes(fill = test_score_n)) +
  scale_fill_gradient(low = "green", high = "red")+ 
  stat_contour(bins = 5) +
  xlab("education") +
  ylab("effort") +
  guides(fill = guide_colorbar(title = "lab")) 

##### P3
df_grid <- expand.grid(education=seq(0,100, by=1), effort=seq(0,100,by=1))
df_grid <- df_grid %>% 
  mutate(test_score = 2*education + 4*effort + 0*education*effort,
         # test_score_n is test score with noise. 
         test_score_n = test_score + rnorm(n = nrow(df_grid), mean = 0, sd = sd(test_score)/10 ) ) %>%
  mutate(test_score_n = ntile(test_score_n, 100))


p3 <- ggplot(df_grid, aes(x = education, y = effort, z = test_score_n)) +
  # stat_contour(geom = "polygon", aes(fill = ..level..)) +
  geom_tile(aes(fill = test_score_n)) +
  scale_fill_gradient(low = "green", high = "red")+ 
  stat_contour(bins = 5) +
  xlab("education") +
  ylab("effort") +
  guides(fill = guide_colorbar(title = "lab")) 


####
grid.arrange(p1, p2, p3, nrow = 3)

```


# Example

We should include the interaction effect in the regression since the interaction terms is statistically significant. And including the interaction term, increase the adjusted $R^2$.

```{r}
# load library
library(dplyr)
library(ggplot2)

# simulate data
N_obs = 1000
set.seed(1)
df_master <- data.frame(education = runif(n = N_obs, min = 0, max = 100),
                        effort = runif(n = N_obs, min = 0, max = 100) )

# generate test score with noise
df_master <- df_master %>% 
  mutate(test_score = 2*education + 4*effort + 0.2*education*effort,
         # test_score_n is test score with noise. 
         test_score_n = test_score + rnorm(n = N_obs, mean = 0, sd = sd(test_score)/10 ) )  

# pairs
pairs(df_master %>% select(-c(test_score)))

# with interaction
summary(lm(test_score_n ~ education * effort, data = df_master   ) )

# without interaction
summary(lm(test_score_n ~ education + effort, data = df_master   ) )

```



# The message

**Suppose our education level is high, a unit increase in effort lead to a higher increase in income, compared to the case where our education level is low.**

The picture below show that to increase the test score from region 1 to region 2, the increase in effort needed depends on the education level.

```{r}
df_grid <- expand.grid(education=seq(0,100, by=1), effort=seq(0,100,by=1))
df_grid <- df_grid %>% 
  mutate(test_score = 2*education + 4*effort + 0.2*education*effort,
         # test_score_n is test score with noise. 
         test_score_n = test_score + rnorm(n = nrow(df_grid), mean = 0, sd = sd(test_score)/10 ) )  



ggplot(df_grid, aes(x = education, y = effort, z = test_score_n)) +
  # stat_contour(geom = "polygon", aes(fill = ..level..)) +
  geom_tile(aes(fill = test_score_n)) +
  scale_fill_gradient(low = "green", high = "red")+ 
  stat_contour(bins = 10) +
  xlab("education") +
  ylab("effort") +
  guides(fill = guide_colorbar(title = "lab")) +
  geom_segment(aes(x = 25, y = 0, xend = 25, yend = 40, colour = "Case 2"), arrow=arrow(), size=1.5) +
  geom_segment(aes(x = 75, y = 0, xend = 75, yend = 15, colour = "Case 1"), arrow=arrow(), size=1.5) 


```

